---
title: "Creating .shp-derived species range maps using R"
author: "MA Hirst, Ph.D."
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is similar to the code I used to create species distribution maps with sampling sites, a legend, and a scale bar in [this journal article](https://onlinelibrary.wiley.com/doi/full/10.1002/ajp.23538).

You can get permission to download specific .shp files (shapefiles) for species from the [IUCN Red List of Threatened Species](https://www.iucnredlist.org).

Once you have your .shp file, navigate to the directory where it is located. Then install and load these libraries:

```{r libs, message=FALSE, warning=FALSE}
library(maps)
library(mapdata)
library(maptools)
library(sp)
library(scales)
library(tmap)
library(raster)
library(ggplot2) 
```

## Create your map

I already know which countries these species are found in, and my sampling sites are restricted to Mexico, so I will focus on Mexico and adjacent countries. For this map, I used shapefile data for two species, *Alouatta palliata* and *Alouatta pigra*. I include basic legends and a scale bar.


```{r map, echo=TRUE, message=FALSE, warning=FALSE}

# base map
map('worldHires', c('Mexico', 'Guatemala', 'Belize'), col='gray95', fill=T, xlim=c(-95.5,-84.5), ylim=c(16.5,21.5))

# make a shape for the range of A. palliata (apa) and plot it
apashp <- readShapePoly('palliata_range/data_0.shp')
plot(apashp, col=alpha('gray', 0.8), border=F, add=TRUE)

# make a shape for the range of A. pigra (api) and plot it
apishp <- readShapePoly('pigra_range/data_0.shp')
plot(apishp, col=alpha('gray25', 0.8), border=F, add=TRUE)

# read in file that has GPS coordinates of sampling sites
p <- read.csv('MapApaApi.csv', header=TRUE) 
points(p$Longitude[1:8], p$Latitude[1:8], pch=19, col='cyan', cex=0.8)
points(p$Longitude[9:13], p$Latitude[9:13], pch=19, col='coral4', cex=0.8)

# define your sampling sites
sites<-c(expression(italic('A. palliata')), expression(italic('A. pigra')))
c<-c('cyan','coral4')

# add legends
legend('left', legend=c(expression(italic('A. palliata')), expression(italic('A. pigra'))), pch=15, bty='n', col=c('gray','gray25'), cex=1.0)
legend('topleft', legend=sites, pch=19, bty='n', col=c('cyan','coral4'), cex=1.0)

# add scalebar in kilometers
scalebar(d=100, xy= c(-95.4, 16.8), type='bar', divs=4, below= "km", lonlat=T, lwd=0.8, cex=0.7)
```
